Simple repression draft

### 2020-06-06 doc/appendix_02_bursty.tex
Implemented most of GC's edits plus more of my own

The objective of this section is to write down the steady-state mRNA
distribution for model 5 in Figure~\ref{fig2:constit_cartoons}. Our claim is
that this model is rich enough that it can capture the expression pattern of
bacteria constitutive promoters. Figure~\ref{figS1:bursty_one_state} shows two
different schematic representations of the model.
Figure~\ref{figS1:bursty_one_state}(A) shows the promoter cartoon model with
burst initiation rate $k_i$, mRNA degradation rate $\gamma$, and mean burst size
$b$. For our derivation of the chemical master equation we will focus more on
Figure~\ref{figS1:bursty_one_state}(B). This representation is intended to
highlight that bursty gene expression allows transitions between mRNA count $m$
and $m'$ even with $m - m' > 1$.

@@

To clean up Equation~\ref{eq:si_master_deltat} we can send the first term on the
right hand side to the left, and divide both sides by $\Delta t$. Upon taking
the limit where $\Delta t \rightarrow 0$ we can write
\begin{equation}
{d \over dt}p(m, t) = (m + 1) \gamma p(m + 1, t)
- m \gamma p(m, t)
+ k_i \sum_{m'=0}^{m-1} G_{m-m'} p(m', t) 
- k_i \sum_{m'=m + 1}^{\infty} G_{m'-m} p(m, t).
\end{equation}
Furthermore, given that the timescale for this equation is set by the mRNA
degradation rate $\gamma$ we can divide both sides by this rate, obtaining
\begin{equation}
{d \over d\tau}p(m, \tau) = (m + 1) p(m + 1, \tau)
- m p(m, \tau)
+ \lambda \sum_{m'=0}^{m-1} G_{m-m'} p(m', \tau) 
- \lambda \sum_{m'=m + 1}^{\infty} G_{m'-m} p(m, \tau),
\label{eq:si_master_ode}
\end{equation}
where we defined $\tau \equiv t \times \gamma$, and $\lambda \equiv k_i/\gamma$.
The last term in Eq.~\ref{eq:si_master_ode} sums all burst sizes except for a 
burst of size zero. We can re-index the sum to include this term, obtaining
\begin{equation}
\lambda \sum_{m'=m + 1}^{\infty} G_{m'-m} p(m, \tau) = \lambda p(m, t) \left[
\underbrace{\sum_{m'=\textcolor{red}{m}}^{\infty}G_{m'-m}}
_{\text{re-index sum to include burst size zero}} -
\underbrace{G_0}_{\text{subtract extra added term}}\right].
\end{equation}
Given the normalization constraint of the geometric distribution, adding the
probability of all possible burst sizes -- including size zero since we
re-indexed the sum -- allows us to write
\begin{equation}
\sum_{m'=m}^{\infty}G_{m'-m} - G_0 = 1 - G_0.
\end{equation}
Substituting this into Eq.~\ref{eq:si_master_ode} results in
\begin{equation}
{d \over d\tau}p(m, \tau) = (m + 1) p(m + 1, \tau)
- m p(m, \tau)
+ \lambda \sum_{m'=0}^{m-1} G_{m-m'} p(m', \tau) 
- \lambda p(m, \tau) \left[ 1 - G_0 \right].
\label{eq:si_master_ode_2}
\end{equation}
To finally get at a more compact version of the equation notice that the third
term in Eq.~\ref{eq:si_master_ode_2} includes burst from size $m'-m = 1$ to size
$m' - m = m$. We can include the term $p(m, t) G_0$ in the sum which allows
bursts of size $m' - m = 0$. This results in our final form for the chemical
master equation
\begin{align}
{d \over d\tau}p(m, \tau) = 
(m + 1) p(m+1, \tau)
- m p(m, \tau) - 
\lambda p(m, \tau)
+ \lambda \sum_{m^\prime=0}^m G_{m-m^\prime} p(m^\prime, \tau).
\label{eq:si_master_unreg}
\end{align}

@@

\begin{figure}[h!]
\centering
\includegraphics{../figures/si/figS0X_reindex_sum.pdf}
\caption{\textbf{Reindexing double sum.} Schematic for how to proceed with the
reindexing of the sum $\sum_{m=0}^\infty \sum_{m'=0}^m$. The trick consists on
noticing that this double sum adds up all of the $(m, m')$ pairs in the lower
triangle part of the $m, m'$ plane.}
\label{figS2:sum_reindex}
\end{figure}

>>>>

The objective of this section is to write down the steady-state mRNA
distribution for model 5 in Figure~\ref{fig2:constit_cartoons}. Our claim is
that this model is rich enough that it can capture the expression pattern of
bacterial constitutive promoters. Figure~\ref{figS1:bursty_one_state} shows two
different schematic representations of the model.
Figure~\ref{figS1:bursty_one_state}(A) shows the promoter cartoon model with
burst initiation rate $k_i$, mRNA degradation rate $\gamma$, and mean burst size
$b$. For our derivation of the chemical master equation we will focus more on
Figure~\ref{figS1:bursty_one_state}(B). This representation is intended to
highlight that bursty gene expression allows transitions between mRNA count $m$
and $m'$ even with $m - m' > 1$.

@@

To clean up Equation~\ref{eq:si_master_deltat} we can send the first term on the
right hand side to the left, and divide both sides by $\Delta t$. Upon taking
the limit where $\Delta t \rightarrow 0$ we can write
\begin{equation}
{d \over dt}p(m, t) = (m + 1) \gamma p(m + 1, t)
- m \gamma p(m, t)
+ k_i \sum_{m'=0}^{m-1} G_{m-m'} p(m', t) 
- k_i \sum_{m'=m + 1}^{\infty} G_{m'-m} p(m, t).
\end{equation}
Furthermore, given that the timescale for this equation is set by the mRNA
degradation rate $\gamma$ we can divide both sides by this rate, obtaining
\begin{equation}
{d \over d\tau}p(m, \tau) = (m + 1) p(m + 1, \tau)
- m p(m, \tau)
+ \lambda \sum_{m'=0}^{m-1} G_{m-m'} p(m', \tau) 
- \lambda \sum_{m'=m + 1}^{\infty} G_{m'-m} p(m, \tau),
\label{eq:si_master_ode}
\end{equation}
where we defined $\tau \equiv t \times \gamma$, and $\lambda \equiv k_i/\gamma$.
The last term in Eq.~\ref{eq:si_master_ode} sums all burst sizes except for a 
burst of size zero. We can re-index the sum to include this term, obtaining
\begin{equation}
\lambda \sum_{m'=m + 1}^{\infty} G_{m'-m} p(m, \tau) = \lambda p(m, t) \left[
\underbrace{\sum_{m'={m}}^{\infty}G_{m'-m}}
_{\text{re-index sum to include burst size zero}} -
\underbrace{G_0}_{\text{subtract extra added term}}\right].
\end{equation}
Given the normalization constraint of the geometric distribution, adding the
probability of all possible burst sizes -- including size zero since we
re-indexed the sum -- allows us to write
\begin{equation}
\sum_{m'=m}^{\infty}G_{m'-m} - G_0 = 1 - G_0.
\end{equation}
Substituting this into Eq.~\ref{eq:si_master_ode} results in
\begin{equation}
{d \over d\tau}p(m, \tau) = (m + 1) p(m + 1, \tau)
- m p(m, \tau)
+ \lambda \sum_{m'=0}^{m-1} G_{m-m'} p(m', \tau) 
- \lambda p(m, \tau) \left[ 1 - G_0 \right].
\label{eq:si_master_ode_2}
\end{equation}
To finally get at a more compact version of the equation notice that the third
term in Eq.~\ref{eq:si_master_ode_2} includes burst from size $m'-m = 1$ to size
$m' - m = m$. We can include the term $p(m, t) G_0$ in the sum which allows
bursts of size $m' - m = 0$. This results in our final form for the chemical
master equation
\begin{align}
{d \over d\tau}p(m, \tau) = 
(m + 1) p(m+1, \tau)
- m p(m, \tau) - 
\lambda p(m, \tau)
+ \lambda \sum_{m^\prime=0}^m G_{m-m^\prime} p(m^\prime, \tau).
\label{eq:si_master_unreg}
\end{align}

@@

\begin{figure}[h!]
\centering
\includegraphics[width=5cm]{../figures/si/figS0X_reindex_sum.pdf}
\caption{\textbf{Reindexing double sum.} Schematic for reindexing the sum
$\sum_{m=0}^\infty \sum_{m'=0}^m$. Blue circles depict the 2D grid of
nonnegative integers restricted to the lower triangular part of the $m, m'$
plane. The trick is that this double sum runs over all $(m, m')$ pairs with
$m'\le m$. Summing $m$ first instead of $m'$ requires determining the
boundary: the upper boundary of the $m'$-first double sum becomes the
lower boundary of the $m$-first double sum.}
\label{figS2:sum_reindex}
\end{figure}

### 2020-06-06 doc/appendix_03_bayesian_inference.tex
Implemented most of GC's edits plus more of my own

\subsubsection{Bayes theorem}

Bayes theorem is a simple mathematical statement that can apply to \textit{any}
logical conjecture. For two particular events $A$ and $B$ that potentially 
depend on each other Bayes theorem gives us a recipe for how to update our 
beliefs about one, let us say $B$, given some state of knowledge, or lack thereof, about
$A$. In its most classic form Bayes theorem is written as
\begin{equation}
P(B \mid A) = {P(A \mid B) P(B) \over P(A)},
\end{equation}
where the vertical line $\mid$ is read as ``given that''. So $P(B \mid A)$ is
read as probability of $B$ given that $A$ took place. $A$ and $B$ can be any
logical assertion. In particular the problem of Bayesian inference focuses on
the question of finding the probability distribution of a particular parameter
value given the data.

For a given model with a set of parameters $\vec{\theta} = (\theta_1, \theta_2,
\ldots, \theta_n)$, the so-called \textit{posterior distribution} 
$P(\vec{\theta} \mid D)$, where $D$ is the experimental data, quantifies the
plausibility of a set of parameter values given our observation of some
particular dataset. In other words, through the application of Bayes formula we
update our beliefs on the possible values that parameters can take upon learning
the outcome of a particular experiment. We specify the word ``update'' as we
come to every inference problem with prior information about the plausibility of
particular regions of parameter space even before performing any experiment.
Even when we claim as researchers that we are totally ignorant about the values
that the parameters in our models can take, we always come to a problem with
domain expertise that can be exploited. If this was not the case, it is likely
that the formulation of our model is not going to capture the phenomena we claim
to want to understand. This prior information is captured in the \textit{prior
probability} $P(\vec{\theta})$. The relationship between how parameter values
can connect with the data is enconded in the \textit{likelihood function} $P(D
\mid \vec{\theta})$. Our theoretical model, whether deterministic or
probabilistic, is encoded in this term that can be intuitively understood as the
probability of having observed the particular experimental data we have at hand
given that our model is parametrized with the concrete values $\vec{\theta}$. 
Implicitly here we are also conditioning on the fact that our theoretical model
is ``true,'' i.e. the model itself if evaluated or simulated in the computer is
capable of generating equivalent datasets to the one we got to observe in an 
experiment. In this way Bayesian inference consists of applying Bayes formula 
as 
\begin{equation}
P(\vec{\theta} \mid D) \propto P(D \mid \vec{\theta}) P(\vec{\theta}).
\end{equation}
Notice than rather than writing the full form of Bayes theorem, we limit 
ourselves to the terms that depend on our quantity of interest -- that is the 
parameter values themselves $\vec{\theta}$ -- as the denominator $P(D)$ only
serves as a normalization constant.

@@

\subsubsection{Prior selection}
The different models explored in this work embraced different levels of
coarse-graining that resulted in a diverse number of parameters for different
models. For each of these model configurations Bayes theorem demands from us to
represent our preconceptions on the possible parameter values in the form of the
prior $P(\vec{\theta})$. Throughout this work for models with $> 1$ parameter we
assign independent priors to each of the parameters; this is
\begin{equation}
P(\vec{\theta}) = \prod_{i=1}^n P(\theta_i).
\end{equation}
Although it is common practice to use non-informative, or maximally
uninformative priors, we are of the mindset that this is a disservice to the
philosophical and practical implications of Bayes theorem. It sounds almost
contradictory to claim that can we represent our thinking about a natural
phenomena in the form of a mathematical model -- which itself is a bold claim of
deeply understanding the features that matter in the object of study -- but we
have absolutely no idea what the parameter values ought to be. We therefore make
use of our own expertise, many times in the form of order-of-magnitude
estimates, to write down prior distributions for our parameters. In summary then
we don’t want to be maximally uninformative because we aren’t completely
uninformed.

For our particular case all of the datasets from~\cite{Jones2014} used in this
paper have $\mathcal{O}(10^3)$ data points. What this implies is that our
particular choice of priors will not significantly affect our inference as long
as they are broad enough. A way to see why this is the case is to simply look at
Bayes theorem. For $N ~ 1000-3000$ datum all of the independent of each other
and $n \ll 10^3$ parameters Bayes theorem reads as
\begin{equation}
P(\vec{\theta} \mid D) \propto \prod_{k=1}^{N} P(d_k \mid \vec{\theta})
\prod_{i=1}^n P(\theta_i),
\end{equation}
where $d_k$ represents the $k$-th datum. That means that if our priors span a
wide range of parameter space, the posterior distribution would be dominated by
the likelihood function.

\subsubsection{Expectations and marginalizations}
One of the great powers of Bayesian analysis is \textit{marginalization}.
For models with more than one or two parameters, it is generally unwise
to attempt to draw inferences directly from the full joint posterior
distribution $P(\vec{\theta} \mid D)$...

\mmnote{
\begin{itemize}
% \item Bayes theorem.
% Elaborate on posterior as plausibility over
% space of model parameters.
% \item Choosing likelihoods.
% Easy here, just the steady-state distributions of our master
% equations, assuming iid single-cell measurements.
% \item Choosing priors.
% Not critical here because we are data-rich and nonhierarchical,
% as long as we don't exclude ground truth. In other applications,
% prior can matter a lot. This is a feature, not a bug: it's
% revealing hidden structure that was implied by your modeling
% assumptions and you just didn't realize it. Predictive checks to
% verify you're not excluding any plausible values.
\item Marginalization/expectations.
Analytically hard, trivial with sampling.
\item MCMC.
Lot's of fun and deep analogies to stat mech and
Hamiltonian mechanics we'll allude to but not explore here.
The idea of MCMC: in a perfect world, you'd
wave a magic wand and generate independent draws from target
posterior. But we don't know how. Next best: generate a Markov
chain of $N$ \textit{correlated} samples, estimate an
autocorrelation time $\tau$, then you've got roughly $N/\tau$
effective samples. Obviously with some assumptions.
\item Posterior predictive checks.
So the posterior is fine, but is the model actually not wrong?
PPC is essential to verify that your model is
not blatantly contradicted by the data.
\end{itemize}
}

@@

As for the Poisson promoter model, this expression for the steady-state mRNA
distribution is exactly the likelihood we want to use in Bayes theorem. Again
denoting the single-cell mRNA count data as $D=\{m_1, m_2,\dots, m_N\}$, here
Bayes theorem takes the form
\begin{equation}
p(k_i, b \mid D) \propto p(D\mid k_i,b)p(k_i, b),
\end{equation}
where the likelihood $p(D\mid k_i,b)$ is given by the product of $N$ negative
binomials as in Eq.~\ref{eq:si_neg_bionom}. We only need to choose priors on
$k_i$ and $b$. For the datasets from~\cite{Jones2014} that we are analyzing, as
for the Poisson promoter model above we are still data-rich so the prior's
influence remains weak, but not nearly as weak because the dimensionality of our
model has increased from one to two.

@@

Formally now, denote the set of seven repressor rates to be inferred as
\begin{equation}
\vect{k} =\{k_{Oid}^-, k_{O1}^-, k_{O2}^-,
k_{0.5}^+, k_{1}^+, k_{2}^+, k_{10}^+\}.
\end{equation}
Note that since the repressor copy numbers are not known directly as explained
before, we label their association rates by the concentration of aTc. Bayes
theorem reads simply
\begin{equation}
p(\vect{k}, k_i, b \mid D)
\propto
p(D \mid\vect{k}, k_i, b) p(\vect{k}, k_i, b),
\end{equation}
where $D$ is the set of all $N$ observed single-cell mRNA counts across the
various conditions. We assume that individual single-cell measurements are
independent so that the likelihood factorizes as
\begin{equation}
p(D \mid\vect{k}, k_i, b)
= \prod_{j=1}^N p(m\mid \vect{k}, k_i, b)
= \prod_{j=1}^N p(m\mid k_j^+, k_j^-, k_i, b)
\end{equation}
where $k_j^\pm$ represent the appropriate binding and unbinding rates for the
$j$-th measured cell. Our likelihood function, previously derived in
Appendix~\ref{sec:gen_fcn_appdx}, is given by the rather complicated result in
Eq.~\ref{eq:p_m_bursty+rep_appdx}, which for completeness we reproduce here as
\begin{equation}
\begin{split}
p(m \mid k_R^+, k_R^-, k_i, b) = & ~\frac{
        \Gamma(\alpha + m)\Gamma(\beta + m)\Gamma(k_R^+ + k_R^-)
        }
        {
        \Gamma(\alpha)\Gamma(\beta)\Gamma(k_R^+ + k_R^- + m)
        }
\frac{b^m}{m!}
\\
&\times {_2F_1}(\alpha+m, \beta+m, k_R^++k_R^-+m; -b).
\end{split}
\label{eq:p_m_bursty+rep}
\end{equation}
where $\alpha$ and $\beta$, defined for notational convenience, are
\begin{align}
\begin{split}
\alpha &= \frac{1}{2}
\left(k_i+k_R^-+k_R^+ + \sqrt{(k_i+k_R^-+k_R^+)^2 - 4k_i k_R^-}\right)
\\
\beta &= \frac{1}{2}
\left(k_i+k_R^-+k_R^+ - \sqrt{(k_i+k_R^-+k_R^+)^2 - 4k_i k_R^-}\right).
\end{split}
\end{align}

@@

With priors and likelihood specified we may write down our complete generative model as
\begin{equation}
\begin{split}
\log_{10}k_i &\sim \text{Normal}(0.725, 0.025)\\
\log_{10}b   &\sim \text{Normal}(0.55, 0.025)\\
\log_{10}k_{0.5}^+ &\sim \text{Normal}(-0.45, 0.3)\\
\log_{10}k_{1}^+   &\sim \text{Normal}(0.6, 0.3)\\
\log_{10}k_{2}^+   &\sim \text{Normal}(1.15, 0.3)\\
\log_{10}k_{10}^+  &\sim \text{Normal}(1.5, 0.3)\\
\log_{10}k_{Oid}^- &\sim \text{Normal}(-0.25, 0.3)\\
\log_{10}k_{O1}^-  &\sim \text{Normal}(0.1, 0.3)\\
\log_{10}k_{O2}^-  &\sim \text{Normal}(0.45, 0.3)\\
m &\sim \text{Likelihood}(k_R^+, k_R^-, k_i, b),
\end{split}
\end{equation}
where the likelihood is specified by Eq.~\ref{eq:p_m_bursty+rep}. We ran MCMC
sampling on the full nine dimensional posterior specified by this generative
model.

@@

\mmnote{Outline:
Perhaps urprising that so many model parameters can be inferred from just a few
distributions that, by eye, don't have strong features! Does require some
assumptions, e.g., about equivalence of rates across experiments, but that seems
quite reasonable.
}

>>>>

\subsubsection{Bayes' theorem}

Bayes' theorem is a simple mathematical statement that can apply to \textit{any}
logical conjecture. For two particular events $A$ and $B$ that potentially 
depend on each other Bayes' theorem gives us a recipe for how to update our 
beliefs about one, let us say $B$, given some state of knowledge, or lack thereof, about
$A$. In its most classic form Bayes' theorem is written as
\begin{equation}
P(B \mid A) = {P(A \mid B) P(B) \over P(A)},
\end{equation}
where the vertical line $\mid$ is read as ``given that''. So $P(B \mid A)$ is
read as probability of $B$ given that $A$ took place. $A$ and $B$ can be any
logical assertion. In particular the problem of Bayesian inference focuses on
the question of finding the probability distribution of a particular parameter
value given the data.

For a given model with a set of parameters $\vec{\theta} = (\theta_1, \theta_2,
\ldots, \theta_n)$, the so-called \textit{posterior distribution} 
$P(\vec{\theta} \mid D)$, where $D$ is the experimental data, quantifies the
plausibility of a set of parameter values given our observation of some
particular dataset. In other words, through the application of Bayes' formula we
update our beliefs on the possible values that parameters can take upon learning
the outcome of a particular experiment. We specify the word ``update'' as we
come to every inference problem with prior information about the plausibility of
particular regions of parameter space even before performing any experiment.
Even when we claim as researchers that we are totally ignorant about the values
that the parameters in our models can take, we always come to a problem with
domain expertise that can be exploited. If this was not the case, it is likely
that the formulation of our model is not going to capture the phenomena we claim
to want to understand. This prior information is captured in the \textit{prior
probability} $P(\vec{\theta})$. The relationship between how parameter values
can connect with the data is enconded in the \textit{likelihood function} $P(D
\mid \vec{\theta})$. Our theoretical model, whether deterministic or
probabilistic, is encoded in this term that can be intuitively understood as the
probability of having observed the particular experimental data we have at hand
given that our model is parametrized with the concrete values $\vec{\theta}$. 
Implicitly here we are also conditioning on the fact that our theoretical model
is ``true,'' i.e. the model itself if evaluated or simulated in the computer is
capable of generating equivalent datasets to the one we got to observe in an 
experiment. In this way Bayesian inference consists of applying Bayes' formula 
as 
\begin{equation}
P(\vec{\theta} \mid D) \propto P(D \mid \vec{\theta}) P(\vec{\theta}).
\end{equation}
Notice than rather than writing the full form of Bayes' theorem, we limit 
ourselves to the terms that depend on our quantity of interest -- that is the 
parameter values themselves $\vec{\theta}$ -- as the denominator $P(D)$ only
serves as a normalization constant.

@@

\subsubsection{Prior selection}
The different models explored in this work embraced different levels of
coarse-graining that resulted in a diverse number of parameters for different
models. For each of these model configurations Bayes' theorem demands from us to
represent our preconceptions on the possible parameter values in the form of the
prior $P(\vec{\theta})$. Throughout this work for models with $> 1$ parameter we
assign independent priors to each of the parameters; this is
\begin{equation}
P(\vec{\theta}) = \prod_{i=1}^n P(\theta_i).
\end{equation}
Although it is not uncommon practice to use non-informative, or maximally
uninformative priors, we are of the mindset that this is a disservice to the
philosophical and practical implications of Bayes' theorem. It sounds almost
contradictory to claim that can we represent our thinking about a natural
phenomenon in the form of a mathematical model -- in the context of Bayesian
inference this means choosing a form for the likehihoods, and even making
this choice presupposes prior understanding or assumptions as to the
relevant features in the system under study -- but that we have absolutely
no idea what the parameter values could or could not be. We therefore make
use of our own expertise, many times in the form of order-of-magnitude
estimates, to write down weakly-informative prior distributions
for our parameters.

For our particular case all of the datasets from~\cite{Jones2014} used in this
paper have $\mathcal{O}(10^3)$ data points. What this implies is that our
particular choice of priors will not significantly affect our inference as long
as they are broad enough. A way to see why this is the case is to simply look at
Bayes' theorem. For $N ~ 1000-3000$ datum all of the independent of each other
and $n \ll 10^3$ parameters Bayes' theorem reads as
\begin{equation}
P(\vec{\theta} \mid D) \propto \prod_{k=1}^{N} P(d_k \mid \vec{\theta})
\prod_{i=1}^n P(\theta_i),
\end{equation}
where $d_k$ represents the $k$-th datum. That means that if our priors span a
wide range of parameter space, the posterior distribution would be dominated by
the likelihood function.

\subsubsection{Expectations and marginalizations}
For models with more than one or two parameters, it is generally difficult
to visualize or reason about the full joint posterior distribution
$P(\vec{\theta} \mid D)$ directly.
One of the great powers of Bayesian analysis is \textit{marginalization},
allowing us to reduce the dimensionality to only the parameters of
immediate interest by averaging over the other dimensions.
Formally, for a three dimensional model with parameters
$\theta_1$, $\theta_2$, and $\theta_3$, we can for instance
marginalize away $\theta_3$ to produce a 2D posterior as
\begin{equation}
P(\theta_1, \theta_2 \mid D) \propto
        \int_{\theta_3} d\theta_3 \,P(\theta_1, \theta_2, \theta_3 \mid D),
\end{equation}
or we can marginalize away $\theta_1$ and $\theta_3$ to produce the
1D marginal posterior of $\theta_2$ alone, which would be
\begin{equation}
P(\theta_2 \mid D) \propto
        \int_{\theta_1} d\theta_1 \int_{\theta_3} d\theta_3
        \,P(\theta_1, \theta_2, \theta_3 \mid D).
\end{equation}
Conceptually, this is what we did in generating the 2D slices of the
full 9D model in Figure~\ref{fig4:repressed_post_full}(A).
In practice, this marginalization is even easier with Markov Chain Monte Carlo
samples in hand. Since each point is simply a list of parameter values,
we simply ignore the parameters which we want to marginalize
away~\cite{Gelman2013}.

\subsubsection{Markov Chain Monte Carlo}
The theory and practice of Bayesian inference with Markov Chain Monte Carlo
(MCMC) is a rich subject with fascinating and deep analogies to statistical
mechanics, even drawing on classical Hamiltonian mechanics and general
relativity in its modern incarnations. We refer the interested reader
to~\cite{Gelman2013} and~\cite{Betancourt2018} for excellent introductions.
Here we merely give a brief summary of
the MCMC computations carried out in this work.

We used the Python package \texttt{emcee} for most of the MCMC sampling
in this work. For the constitutive promoter inference, we also ran
sampling with the excellent Stan modeling language as a check. We did not
use Stan for the inference of the simple repression model because
implementing the gradients of the hypergeometric function ${_2F_1}$
appearing in Eq.~\ref{eq:p_m_bursty+rep_appdx}, the probability
distribution for our bursty model with repression, would have been
an immensely challenging task. \texttt{emcee} was more than adequate for
our purposes, and we were perhaps lucky that the 9-D posterior model
for the model of simple repression with bursty promoter was quite well
behaved and did not require the extra power of the Hamiltonian Monte
Carlo algorithm provided by Stan~\cite{Carpenter2017}.
Source code for all statistical inference will be made available at
\url{https://github.com/RPGroup-PBoC/bursty_transcription}.

@@

As for the Poisson promoter model, this expression for the steady-state mRNA
distribution is exactly the likelihood we want to use in Bayes' theorem. Again
denoting the single-cell mRNA count data as $D=\{m_1, m_2,\dots, m_N\}$, here
Bayes' theorem takes the form
\begin{equation}
p(k_i, b \mid D) \propto p(D\mid k_i,b)p(k_i, b),
\end{equation}
where the likelihood $p(D\mid k_i,b)$ is given by the product of $N$ negative
binomials as in Eq.~\ref{eq:si_neg_bionom}. We only need to choose priors on
$k_i$ and $b$. For the datasets from~\cite{Jones2014} that we are analyzing, as
for the Poisson promoter model above we are still data-rich so the prior's
influence remains weak, but not nearly as weak because the dimensionality of our
model has increased from one to two.

@@

Formally now, denote the set of seven repressor rates to be inferred as
\begin{equation}
\vect{k} =\{k_{Oid}^-, k_{O1}^-, k_{O2}^-,
k_{0.5}^+, k_{1}^+, k_{2}^+, k_{10}^+\}.
\end{equation}
Note that since the repressor copy numbers are not known directly as explained
before, we label their association rates by the concentration of aTc. Bayes
theorem reads simply
\begin{equation}
p(\vect{k}, k_i, b \mid D)
\propto
p(D \mid\vect{k}, k_i, b) p(\vect{k}, k_i, b),
\end{equation}
where $D$ is the set of all $N$ observed single-cell mRNA counts across the
various conditions. We assume that individual single-cell measurements are
independent so that the likelihood factorizes as
\begin{equation}
p(D \mid\vect{k}, k_i, b)
= \prod_{j=1}^N p(m\mid \vect{k}, k_i, b)
= \prod_{j=1}^N p(m\mid k_j^+, k_j^-, k_i, b)
\end{equation}
where $k_j^\pm$ represent the appropriate binding and unbinding rates for the
$j$-th measured cell. Our likelihood function, previously derived in
Appendix~\ref{sec:gen_fcn_appdx}, is given by the rather complicated result in
Eq.~\ref{eq:p_m_bursty+rep_appdx}, which for completeness we reproduce here as
\begin{equation}
\begin{split}
p(m \mid k_R^+, k_R^-, k_i, b) = & ~\frac{
        \Gamma(\alpha + m)\Gamma(\beta + m)\Gamma(k_R^+ + k_R^-)
        }
        {
        \Gamma(\alpha)\Gamma(\beta)\Gamma(k_R^+ + k_R^- + m)
        }
\frac{b^m}{m!}
\\
&\times {_2F_1}(\alpha+m, \beta+m, k_R^++k_R^-+m; -b).
\end{split}
\label{eq:p_m_bursty+rep_infreprint}
\end{equation}
where $\alpha$ and $\beta$, defined for notational convenience, are
\begin{align}
\begin{split}
\alpha &= \frac{1}{2}
\left(k_i+k_R^-+k_R^+ + \sqrt{(k_i+k_R^-+k_R^+)^2 - 4k_i k_R^-}\right)
\\
\beta &= \frac{1}{2}
\left(k_i+k_R^-+k_R^+ - \sqrt{(k_i+k_R^-+k_R^+)^2 - 4k_i k_R^-}\right).
\end{split}
\end{align}

@@

With priors and likelihood specified we may write down our complete generative model as
\begin{equation}
\begin{split}
\log_{10}k_i &\sim \text{Normal}(0.725, 0.025)\\
\log_{10}b   &\sim \text{Normal}(0.55, 0.025)\\
\log_{10}k_{0.5}^+ &\sim \text{Normal}(-0.45, 0.3)\\
\log_{10}k_{1}^+   &\sim \text{Normal}(0.6, 0.3)\\
\log_{10}k_{2}^+   &\sim \text{Normal}(1.15, 0.3)\\
\log_{10}k_{10}^+  &\sim \text{Normal}(1.5, 0.3)\\
\log_{10}k_{Oid}^- &\sim \text{Normal}(-0.25, 0.3)\\
\log_{10}k_{O1}^-  &\sim \text{Normal}(0.1, 0.3)\\
\log_{10}k_{O2}^-  &\sim \text{Normal}(0.45, 0.3)\\
m &\sim \text{Likelihood}(k_R^+, k_R^-, k_i, b),
\end{split}
\end{equation}
where the likelihood is specified by Eq.~\ref{eq:p_m_bursty+rep_infreprint}.
We ran MCMC sampling on the full nine dimensional posterior specified
by this generative model.

@@
